---
title: "Alternative Fitting Approaches to Linear Models"
author: "Nitant Agarwal"
date: "10 November 2017"
output: html_document
---


## Alternative Fitting Approaches to Linear Models

Find alternative fitting approaches to the normal linear models besides the Ordinary Least Squares(OLS). OLS is ideal when the underlying relationship is Linear and we have n>>p. But if n is n is not much greater than p, there can be a lot of variablitiy in the fit which can result in either overfitting and very poor predictive ablity. 


### Model Interpretability:

In multiple regression, we have the problem of using many predictors which don't add much to the predictive ability or are highly correlated or cause unncessary complexity in the model. So we propose the following methods to improve the model interpretability

###Method 1: Subset Selection

By selecting a subset of p predictors using different techniques which we believe has a concrete influence on the response, we try to fit the model using the simple least squares method.

###Method 2: Shrinkage

By constraining or shrinking the estimated coefficients, we can sucessfully reduce the variance associated with the model with a little increase in bias which improves the predictive accuracy.

###Method 3: Dimension Reduction

We can also use dimension reduction through which we can reduce the correlated variables. This method typically involved proejecting the predictors into a multidimensional subspace and computing different combinators or projections of the bbariables and these projections are used as predictors in the model.


```{r echo=FALSE}
library(MASS)
library(e1071)
library(caret)
library(ggplot2)
library(dplyr)
library(knitr)
library(tidyverse)
library(knitr)
library(reshape2)
library(leaps)
library(boot)
library(plyr)
library(lars)
library(elasticnet)
library(GGally)
```
We are dividing the data into 75% training and 25% testing data. 

```{r}
set.seed(1050)
subset_data <- sample(nrow(Boston), nrow(Boston)*0.75)
train_data <- Boston[subset_data,]
test_data <- Boston[-subset_data,]
```
#Exploratory Analysis

We perform basic exploratory analysis to understand the underlying variables and their dependancies. 

```{r}
kable(str(Boston), "html")
```
#Box Plots

The box plot indicate strong outliers in the dependant vairbale mdev and also in independant variables such as crim, chas, black, rm and zn. They also indicate that almost none of the variables are normally distributed.

```{r}
Boston %>%
  gather(key = "attribute", value = "value") %>%
  ggplot() +
  geom_boxplot(mapping = aes(x = attribute, y = value), fill = "#1abc9c") +
  facet_wrap(~ attribute, scales = "free")
```

#Correlation Plots

The Correlation plots and the matrix indicate the relationship of dependant variable mdev and   other independant variables and also of multicollinearity between the independant variables. We find rm, ptratio and lstat might be vary influential in determining mdev. 

```{r}
c <- round(cor(Boston),2)
kable(c)
```

```{r}
pairs <- ggpairs(Boston, lower = list(continuous = wrap("smooth", colour = "turquoise4")),
                diag = list(continuous = wrap("barDiag", fill = "turquoise4"))) + 
         theme(panel.background = element_rect(fill = "gray98"),
                axis.line.x = element_line(colour = "gray"),
                axis.line.y = element_line(colour = "gray"))
pairs
```

#Density Plots and Histograms

The density plots and histogram reiterate that none of the variables have a perfectly normal distribution rm has an almost normal distribution and the dependant variable mdev has a right skewed distribution which can be log transformed to get a better model. 

```{r}
ggplot(data = melt(Boston), aes(x = value)) + 
         stat_density() + 
         facet_wrap(~variable, scales = "free")
```

```{r}
ggplot(data = melt(Boston), aes(x = value)) + 
         geom_histogram() + 
         facet_wrap(~variable, scales = "free")
```

#Model Performance Indicators

We will use the following paramters to explain the model performance and the intrinsic differences in the fitting of various models. We can extract all of these results from the fit statement which has a list of stored values for each model.

###AIC- 
Akaike's Information Criterion offers a relative estimate of the infomration lost wen a given model is used to fit the data. It deals with the trade-off between goodness of fit of the model and the complexity of the model. It is defined for models fit by maximum likelihood function majorly and is proportional to another measure called Mallow's Cp.The lower the AIC and Mallow's Cp, better the model.

###BIC- 
Bayesian Information Criterion/ Schwartz Criterion offers a similar trade-off between goodness of fit and complexity of model but penalizes the complexity more than AIC as the number of parameters added to the model increases, typically having BIC values > AIC values and smaller models compared to AIC. Lower the BIC, Better the model.

###MSE- 
Mean Square Error is the average distance between the observed values and the predicted values. Lower the MSE, more accurate the model.

#Variable Selection/Regularization

As the number of predictors increaeses, we need to work with constraints or regularization to reduce the number of predictors by order of their importance or predictive ability. We have few commonly used methods and few rarely used methods due to complexity listed below:

#Subset Selection Methods - Discrete Elimination, High Variance

##1. Best Subsets Selection- EXHAUSTIVE ALGORITHM

This method typically finds the best subset of each size k which fits the model the best. It is typically feasible for small number of predictors (less than 30) as 2^p models exists where p is the number of predictors. It fits all models with k predictors where k is the size and selects a single best model of each size based on cross-validated prediction error. This method is highly computationally inefficient but provides the best model. So we have alternate methods.

##2.Forward Selection

This method uses a null model and builds up with one variable at a time until all the predictors are added to the model. The order of addition depends on which predictor provided the best improvement to the fit until addition of extra variables will not guarantee any improvement to the model. This fits much lesser model when compared to best subset method. Forward can be applied even in conditions where p>n while other methods cannot be used.

##3.Backward Selection

This method starts with a full model and starts reducing the predictors one at a time based on the insignificance of that variable to the entire process. This cannot be used in conditions where p>n.

##4.Stepwise Selection- GREEDY ALGORITHM

This method uses a mix of forward and backward and has the ability to move in either direction adding one variable at a time till overfitting is detected. This method can also delete a variable which does not add much improvement to the fit and hence move in both direction.. This is a greedy algorithm as the model with 5 variables is not the best model of size 5, like in best subsets.

##Implementation

We start with best subsets method, which tells us that indus, age and black may not be great predictors.

```{r}
#Variable Selection
par(mfrow=c(1,1))
#Best subset selection using BIC
subset_result2 <- regsubsets(medv ~., data = train_data, nbest = 2, nvmax = 14 )
summary(subset_result2)
```

```{r}
plot(subset_result2)
```

We move on to Stepwise, Forward and Backward selection. All of them produce the same model which just removed the indus variable from the model. 

```{r}
nullmodel <- glm(medv ~1, data = train_data)
fullmodel <- glm(medv ~., data = train_data)
model.step <- step(nullmodel, scope = list(lower = nullmodel, upper = fullmodel), direction = "both")

#Forward
model.forward <- step(nullmodel, scope = list(lower = nullmodel, upper = fullmodel), direction = "forward")
summary(model.forward)

#Backward
model.backward <- step(fullmodel, direction = "backward")
summary(model.backward)

```

```{r}
summary(model.step)
```

The final results for the subset selection method are:
In-Sample MSE- 25.04
Out of Sample MSE- 13.23
AIC- 2283.81
BIC- 2375.06

#Shrinkage Methods - Continous elimination, Lower Variance

##1.Ridge Regression

Ridge regression does not select variables but shrinks the correlated predictor coefficient estimates towards each other or zero based on the size of the tuning parameter/complexity paramater. When the tuning parameter is equal to 0, it becomes the OLS process.
Ridge regression also requires the predictors to be standardized (centered) before starting the regression. Ridge Regression shrinks the coefficients of the low-variance components more than the high variance components. While OLS regression estimates have high variability, by scaling, ridge regression acheives much lesser variability and MSE scores but still has all variables in the model.

##Ridge Vs OLS

While OLS regression estimates have high variability, by scaling, ridge regression acheives much lesser variability and MSE scores but still has all variables in the model. As the tuning parameter increases, the variance decresses with little increase in the bias. Regular OLS has no bias but higher variance.

##2.LASSO

LASSO actually performs variable selection along with shrinkage and is prone to produce more interpretable models when compared to Ridge. LASSO also involves penalizing the OLS estimates of the betas but can zero out some of the betas and picks just one prdictor from a group of correlated variables (reducing the number of variables in the model) as it shrinks betas by fixed amounts whereas Ridge regression shrinks everything proportionally. The tuning paramters are determined by Cross Validation.

##LASSO vs Ridge

Lasso performs better where there are small number of predictors with significant coefficients and others are very small or equal to zero whereas Ridge performs better when response is a function of a large number of predictors with coefficients of roughly the same size.

##3. Elastic Net

Elastic Net mixes the penalty of ridge and lasso and is useful when the number of predictors is very large and we want to select more than one predictor from a group of correlated variables. It behaves similar to lasso but removes any degeneracies and wild behaviour caused by extreme correlations.

##Implementation

We start with Ridge Regression without Cross-Validation:

```{r}
ridge <- train(medv~., data = train_data, method = "ridge", lambda = 4, preProcess = c('scale', 'center'))

ridge
```

```{r}
ridge.pred <- predict(ridge, test_data)

mean((ridge.pred - test_data$medv)^2)
```

We perform Cross-Validation, where the tuning parameter was chosen by minimum RMSE. When there the tuning parameter is zero, the error is very high. As the tuning parameter increases, the error decreases but beyond a value, bias becomes larger and results in a under-fit model. 

```{r}
#Cross-Validation to pick Lambda
ridge_grid <- expand.grid(lambda = seq(0,0.1, length = 15))
set.seed(1050)
ridge_model <- train(medv~., data = train_data, method = "ridge", preProcess = c('center', 'scale'), tuneGrid = ridge_grid, trControl = trainControl(method = "cv"))

ridge_model
```

```{r}
ridge.pred <- predict(ridge_model, test_data)

mean((ridge.pred - test_data$medv)^2)
```

```{r}
update(plot(ridge_model), xlab = "Penalty", 
      main = "The Cross-Valiation Profiles for Ridge Regression Model")
```

Without Cross-Validation:
MSE - 16.732
R^2 - 0.68
With Cross Validation:
MSE - 16.71
R^2 - 0.728

We see that Ridge has better MSE than the normal subset selection methods.

We then move on to Lasso Regression:

```{r}
lasso <- train(medv~., train_data, 
               method = "lasso", 
               preProcess = c("center", "scale"))

lasso
```

```{r}
lasso.pred <- predict(lasso, test_data)
mean((lasso.pred - test_data$medv)^2)
```

```{r}
lasso.grid <- expand.grid(fraction =seq(0.05,1, length = 20))
set.seed(1050)
lasso_model <- train(medv~., data = train_data,
                     method = "lasso", 
                     preProcess = c("center", "scale"),
                     tuneGrid = lasso.grid,
                     trControl = trainControl(method = "cv"))

lasso_model
```


```{r}
lasso.pred <- predict(lasso_model, test_data)
mean((lasso.pred - test_data$medv)^2)
```

Without Cross - Validation:
MSE - 16.74
R^2 - 0.68
With Cross - Validation:
MSE - 16.73
R^2 - 0.728

lasso seems to have similar MSE compared to ridge. 

We then move to Elastic Net:

```{r}
e_net <- train(medv~., data = train_data,
               method = "enet",
               preProcess = c("scale", "center"))
e_net
```


```{r}
enet.pred <-  predict(e_net, test_data)
mean((enet.pred - test_data$medv)^2)
```

We then perform Cross - Validation:

```{r}
enet.grid <- expand.grid(lambda =seq(0,0.01,.1), fraction = seq(0.05,1, length = 20))
set.seed(1050)
enet_model <- train(medv~., data = train_data,
                     method = "enet", 
                     preProcess = c("center", "scale"),
                     tuneGrid = enet.grid,
                     trControl = trainControl(method = "cv"))

enet_model
```

```{r}
enet.pred <-  predict(enet_model, test_data)
mean((enet.pred - test_data$medv)^2)
```

```{r}
update(plot(enet_model), main = "The Cross-validation Profiles for Elastic Net Model")
```

Without Cross - Validation:
MSE - 16.95
R^2 - 0.70
With Cross - Validation:
MSE - 16.73
R^2 - 0.728

#Dimension Reduction Methods

Unlike the above methods where we contrlled for variance by subsets or by shrinking coefficients, we will now transform the predictor variables and then fit a least squares model using these transformed variables.

##1. Principal Components Regression

PCA (Principal Component Analysis) is a methodology used to derive a low-dimensional set of features from a very large set of predictors. The first PC (principal component) direction of the data is along which the variance of the observations is the highest. This is the line which fits very closely to the data. The second PC is uncorrelated to first PC (orthogonal to first PC) and has the highest variance subjected to a constraint. We use linear combinations of data in orthogonal directions which maximize the variance captured by our model instead of dismissing a variable out of two correlated variables in OLS.
In Principle Components Regression, we construct M principal components and use these in a linear fashion to obtain the least squares. We fit the model only with those variables which explain most of the variability in the data and the relationship with the response variable, thus reducing the risk of overfitting.

There is no feature selection happening as it is a linear combination of all the p original features. PCR is better when the first few PCs are sufficient to capture most of the variation in the data.

##2. Partial Least Squares

We don't use the response variable to determine the principal component directions, hence making it a unsupervised method. This reduces the guarantee of the principal components explaining the response completely.

Partial Least squares (PLS) method is more supervised dimension reduction technique which identifies a smaller set of features which are linear combinations of original features using the response variable to identify these new features.

The methods places higher weights on variables that are more related to the response variable. To attain the directions, PLS asjusts each of these variables for the first component by regressing them on first component and taking residuals. The residuals are the remaining information which has not been explained by the first PLS direction. This is them iterated M times to identify multiple PLS components.

##Implementation

We start with PCR with Cross-Validation:

```{r}
set.seed(1050)
pcr_model <- train(medv~., data = train_data,
                   method = "pcr",
                   preProcess = c("center", "scale"),
                   tuneGrid = expand.grid(ncomp = 1:13),
                   trControl = trainControl(method = "cv"))
summary(pcr_model)
```

```{r}
pcr_pred <- predict(pcr_model, test_data)
mean((pcr_pred - test_data$medv)^2)
```

```{r}
df_pcr <- data.frame(predicted = pcr_pred, observed = test_data$medv, residual = test_data - pcr_pred)

ggplot(df_pcr, aes(x = predicted, y = observed)) + 
  geom_point() + 
  geom_abline(intercept = 0, slope = 1, colour = "blue") + 
  ggtitle("Principal Component Regression Predicted VS Observed")
```

The following are the results:
With Cross Validation:
MSE - 16.73

##PLS with Cross validation

```{r}
set.seed(1050)
pls_model <- train(medv~., data = train_data, 
                   method = "pls",
                   preProcess = c("scale", "center"),
                   tuneGrid = expand.grid(ncomp = 1:13),
                   trControl = trainControl(method = "cv"))

pls_model
```

```{r}
pls_pred <- predict(pls_model,  test_data)
mean((pls_pred - test_data$medv)^2)
```

```{r}
df_pls <- data.frame(predicted = pls_pred, observed = test_data$medv, residual = test_data - pls_pred)

ggplot(df_pls, aes(x = predicted, y = observed)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, colour = "blue") + 
  ggtitle("Partial Least Squares Predicted VS Observed")
```

```{r}
pcr_model$results$model <- "pcr"
pls_model$results$model <- "pls"

df_pcr_pls <- rbind(pcr_model$results, pls_model$results)
ggplot(df_pcr_pls, aes(x = ncomp, y = RMSE, colour = model)) + 
  geom_line() + 
  geom_point() +
  ggtitle("PCR VS PLS")
```

##Importance of the predictors

```{r}
pls_imp <- varImp(pls_model, scale = F)
plot(pls_imp, scale = list(y = list(cex = .95)))
```

With Cross Validaion:
MSE - 16.78
